---
title: "Using proteus R package: TMT data"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using proteus R package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(proteus)
library(knitr)
library(ggplot2)
options(width = 80)
knitr::opts_chunk$set(echo = TRUE)
```

This tutorial demonstrates how to analyse data from TMT MS/MS experiment in *Proteus*. We strongly suggest the user should read the "unlabelled" tutorial first and familiarize themselves with the package. 

```{r, eval=FALSE}
vignette("unlabelled", package="proteus")
```

Here, we are only going to point out the differences between unlabelled and TMT analysis. For this, we use data from [Crozier et al. 2017](http://www.mcponline.org/content/16/12/2254.long), Pride accession [PXD005968](ttps://www.ebi.ac.uk/pride/archive/projects/PXD005968). These data are distributed in a separate package *proteusTMT*, which needs loading first:

```{r load_data, echo=FALSE}
library(proteusTMT, warn.conflicts=FALSE)
data(proteusTMT)
```

```{r load_data_dummy, eval=FALSE}
library(proteusTMT)
data(proteusTMT)
```

# Input data

## Columns in evidence file

The default `measure.cols` object is designed for unlabelled data. For TMT data we need to specify all reporter intensity columns. In out example we have 10 reporter columns, numbered from 0 to 9:

```{r measure_columns, cache=FALSE}
measCols <- paste0("Reporter intensity ", 0:9)
names(measCols) <- paste0("reporter_", 0:9)
```

The resulting object is

```{r measure_columns_object}
str(as.list(measCols))
```

We are going to use the default `evidenceColumns` object, but the user should carefully check that the column names correspond to those in the evidence file.


## Read evidence file

Due to its large size we do not attach the TMT evidence file to this package. The command to read the file would be:

```{r read_evidence, eval=FALSE}
evi <- readEvidenceFile(evidenceFile, measure.cols=measCols)
```

We do attach an example processed evidence data with the package, it is called `evi`. It contains all reported intensity columns:

```{r head_evidence}
head(evi)
```

## <a id="Metadata"></a>Metadata

Metadata for this experiment needs to specify all the reporter columns, experiments and corresponding condition and replicates:

```{r metadata, eval=FALSE}
metadataFile <- system.file("extdata", "metadata.txt", package="proteusTMT")
meta <- read.delim(metadataFile, header=TRUE, sep="\t")
```

```{r show_metadata}
meta
```

## No 'Experiment' column?

It is possible to run *MaxQuant* for one experiment where conditions are encoded in reporter columns only. In this case the evidence file will not contain the column 'Experiment'. *Proteus* can deal with such configuration. All you need is to remove the 'experiment' field from the default column list:

```{r evidence_without_experiment}
eviCols <- evidenceColumns
eviCols$experiment <- NULL
```

and use argument `data.cols = eviCols` in `readEvidenceFile` function. The 'Experiment' column will be skipped. Obviously, metadata should not contain 'Experiment' column either.

Please note that in such case peptide and protein objects created by *Proteus* will contain a metadata data frame (`metadata` field in the object, see, e.g., `pepdat$metadata`) with column `experiment` filled with dummy values. This is done for easier internal processing. OK, we are lazy and it was easier to do it this way.

# Peptide data

## Create a peptide dataset

We chose the median to aggregate peptide intensities from multiple entries in the evidence data:

```{r make_peptides, eval=FALSE}
pepdat <- makePeptideTable(evi, meta, measure.cols=measCols, aggregate.fun=aggregateMedian, experiment.type="TMT")
```

We can use generic `summary` function to see more information about `xppepdat`.

```{r summary_peptides}
summary(pepdat)
```

## Number of peptides

This is the number of non-zero peptide intensities per sample. 

```{r plot_peptide_count, fig.width=5, fig.height=4}
plotCount(pepdat)
```




# Protein data

## Create protein dataset

We create protein data using the high-flyer method.

```{r make_proteins, eval=FALSE}
prodat <- makeProteinTable(pepdat, aggregate.fun=aggregateHifly, hifly=3)
```

Again, we can use a generic `summary` function to see its properties.

```{r summary_proteins}
summary(prodat)
```

## Normalization

For TMT data we recommend using CONSTANd normalization [Maes et al. 2016](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4974351/pdf/zjw2779.pdf).


```{r normalize_proteins}
prodat.norm <- normalizeTMT(prodat)
```

These two figures show reporter intensity distributions before and after normalization.

```{r, sample_dist, fig.width=7, fig.height=4}
plotSampleDistributions(prodat, fill="replicate") + labs(title="Before")
plotSampleDistributions(prodat.norm, log.scale=FALSE, fill="replicate") + labs(title="After")
```


## Protein clustering

We can use the same function `plotClustering()` to see the dendrogram for the proteins.

```{r plot_clustering_proteins, fig.width=6, fig.height=5}
plotClustering(prodat.norm)
```

# Differential expression

These particular data come from a time-course experiment that would require analysis beyond *Proteus* package. We can however show how differential expression can be used to identify proteins changing between the first and middle time point:

```{r DE_C0_C4}
res <- limmaDE(prodat.norm, conditions=c("C0", "C5"))
```

The distribution of p-values looks fine.

```{r p_value_dist, fig.width=5, fig.height=4}
plotPdist(res)
```

The top of the table of differentially expressed proteins:

```{r de_table_top}
res <- res[order(res$adj.P.Val),]
head(res)
```

And the top protein from the table:

```{r DE_example1, fig.width=8, fig.height=3}
prot <- as.character(res$protein[1])
plotIntensities(prodat.norm, id=prot)
```
