---
title: "Using proteus R package: TMT data"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using proteus R package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{css, echo=FALSE}
#pre code, pre, code {
#  white-space: pre !important;
#  overflow-x: scroll !important;
#  word-break: keep-all !important;
#  word-wrap: initial !important;
#}
```

```{r setup, include=FALSE}
library(proteus)
library(knitr)
library(dplyr)
library(reshape)
library(ggplot2)
library(gplots)
library(grid)
require(gridExtra)
library(dendextend)
options(width = 80)
knitr::opts_chunk$set(echo = TRUE)
```

This tutorial demonstrates how to analyse data from TMT MS/MS experiment in *Proteus*. We strongly suggest the user should read the "unlabelled" tutorial first and familiarise themselves with the package. Here, we are only going to point out the differences between unlaballed and TMT analysis

```{r, eval=FALSE}
vignette("unlabelled", package="proteus")
```

# Input data

## Columns in evidence file

The default `measure.cols` object is designed for unlabelled data. For TMT data we need to specify all reporter intensity columns. In out example we have 10 reporter columns, numbered from 0 to 9:

```{r measure_columns, cache=FALSE}
measCols <- paste0("Reporter intensity ", 0:9)
names(measCols) <- paste0("reporter_", 0:9)
```

The resulting object is

```{r measure_columns_object}
str(as.list(measCols))
```

We are going to use the default `evidenceColumns` object, but the user should carefully check that the column names correspond to those in the evidence gile.


## Read evidence file

Due to its large size we do not attach the TMT evidence file to this package. The command to read the file would be:

```{r read_evidence, eval=FALSE}
evidenceFile <- "../../GreatProt/data/Cell_cycle_TMT/evidence.txt"
xtevi <- readEvidenceFile(evidenceFile, measure.cols=measCols)
```

We do attach an example processed evidence data with the package, it is called `xtevi`. It contains all reported intensity columns:

```{r head_evidence}
head(xtevi)
```

## <a id="Metadata"></a>Metadata

Metadata for this experiment needs to specify all the reporter columns, experiments and corresponding condition and replicates:

```{r metadata, eval=FALSE}
metadataFile <- "../../GreatProt/data/Cell_cycle_TMT/metadata.txt"
xtmeta <- read.delim(metadataFile, header=TRUE, sep="\t")
```

```{r show_metadata}
xtmeta
```


# Peptide data

## Create a peptide dataset

We chose the median to aggregate peptide intensities from multiple entries in the evidence data:

```{r make_peptides, eval=FALSE}
xtpepdat <- makePeptideTable(xtevi, xtmeta, measure.cols=measCols, fun.aggregate=median, experiment.type="TMT")
```

We can use generic `summary` function to see more information about `xppepdat`.

```{r summary_peptides}
summary(xtpepdat)
```

## Number of peptides

This is the number of non-zero peptide intensities per sample. 

```{r plot_peptide_count, fig.width=5, fig.height=4}
plotCount(xtpepdat)
```




# Protein data

## Create protein dataset

We create protein data using the high-flyer method.

```{r make_proteins, eval=FALSE}
xtprodat <- makeProteinTable(xtpepdat, method="hifly", hifly=3)
```

Again, we can use a generic `summary` function to see its properties.

```{r summary_proteins}
summary(xtprodat)
```

## Normalization

For TMT data we recommend using CONSTANd normalization [Maes et al. 2016](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4974351/pdf/zjw2779.pdf).


```{r normalize_proteins}
xtprodat.norm <- normalizeTMT(xtprodat)
```

These two figures show reporter intensity distributions before and after normalization.

```{r, sample_dist, fig.width=7, fig.height=4}
plotSampleDistributions(xtprodat, fill="replicate") + labs(title="Before")
plotSampleDistributions(xtprodat.norm, log.scale=FALSE, fill="replicate") + labs(title="After")
```


## Protein clustering

We can use the same function `plotClustering()` to see the dendrogram for the proteins.

```{r plot_clustering_proteins, fig.width=6, fig.height=5}
plotClustering(xtprodat.norm)
```

# Differential expression

These particular data come from a time-course experiment that would require analysis beyond *Proteus* package. We can however show how differential expression can be used to identify proteins changing between the first and middle time point:

```{r DE_C0_C4}
res <- limmaDE(xtprodat.norm, conditions=c("C0", "C5"))
```

The distribution of p-values looks fine.

```{r p_value_dist, fig.width=5, fig.height=4}
plotPdist(res)
```

The top of the table of differentially expressed proteins:

```{r de_table_top}
res <- res[order(res$adj.P.Val),]
head(res)
```

And the top protein from the table:

```{r DE_example1, fig.width=8, fig.height=3}
prot <- as.character(res$protein[1])
plotProteins(xtprodat.norm, protein=prot)
```
