---
title: "Using proteus R package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using proteus R package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(proteus)
library(knitr)
library(dplyr)
library(reshape)
library(ggplot2)
library(gplots)
library(grid)
require(gridExtra)
library(dendextend)
options(width = 1000)
knitr::opts_chunk$set(echo = TRUE)
```

# Input data

We need two files for proteus. The MaxQuant evidence file and a manually created metadata file. The metadata file is a tab-delimited text file with a header containing several columns of metadata. The mandatory columns are "sample" and "condition".

```{r}
#topDir <- '/Users/mgierlinski/Work/Biology/Paper/GreatProt'
topDir <- '/gpfs/mgierlinski/projects/greatprot_matthias/'
dataDir <- paste0(topDir, '/data')

evidenceFile <- paste0(dataDir, '/1112/evidence.txt')
peptideFile <- paste0(dataDir, '/1112/peptides.txt')
proteinFile <- paste0(dataDir, '/1112/proteinGroups.txt')
metadataFile <- paste0(dataDir, '/m2_metadata.txt')
```

Read evidence file and metadata.

```{r}
evi <- readEvidenceFile(evidenceFile)
meta <- read.delim(metadataFile, header=TRUE, sep="\t")
```

Our recommendation is to clean the evidence file and remove data with number of scans <= 10 and number of points <= 10:

```{r}
evi <- evi[which(evi$numpoints > 10 & evi$numscans > 10),]
```

This is the design of our expeiment:

```{r}
select(meta, sample, condition, replicate)
```

Create a peptide data object.

```{r}
pepdat <- makePeptideTable(evi, meta)
```

It consists of the intensity table

```{r}
head(pepdat$tab)
```

and additional information

```{r}
names(pepdat)
```

Check the total number of peptides and proteins.

```{r}
length(pepdat$peptides)
length(pepdat$proteins)
```
See the number of peptides detected in each sample.

```{r}
plotPeptideCount(pepdat, x.text.size=10)
```

Quick quality control.

```{r}
plotCorrelationMatrix(pepdat)
plotClustering(pepdat)
```

The last sample (BMO-7) looks odd. Lets say we call it "bad" and want to remove from analysis.

```{r}
meta_clean <- meta[which(meta$sample  != 'BMO-7'),]
pepdat <- makePeptideTable(evi, meta_clean)
plotClustering(pepdat)
```

Next, we create a protein data table:

```{r}
prodat <- makeProteinTable(pepdat)
```

Mean-variance relationship.

```{r}
plotMV(prodat)
```

```{r}
plotClustering(prodat)
```

